Generalized Tractability

### 2023-04-20 GeneralizedTractability_230418.tex
Updates from Overleaf

\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

The goal is to find an approximate solution, $\APP_d:\cb_d \times (0,\infty) \to \cg_d$ satisfying
\[
\norm[\cg_d]{\SOL_d(f) - \APP_d(f,\varepsilon)} \le \varepsilon,
\]
where $\cb_d$ is the unit ball in $\cf_d$ and $\APP_d(f,\varepsilon)$ is allowed to depend on arbitrary linear functionals \peter{(note that we also allow adaptive information, i.e., the linear functionals may sequentially depend on each other)}.  In this case, the optimal solution is
\[
\APP_d(f,\varepsilon) = \sum_{i=1}^n \SOL(u_i) \ip[\cf_d]{f}{u_i},
\]
and the \peter{information} complexity of our linear problem is
\begin{equation}\label{eq:comp_def}
\comp(\varepsilon, d) = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\}.
\end{equation}

By definition, the complexity is non-decreasing  as $\varepsilon$ tends to zero.  We also expect it to be non-decreasing as $d$ tends to infinity.  \peter{This means that we are actually considering 
a whole sequence of problems  $\SOL_d : \cf_d \to \cg_d$, for $d\in\naturals$, and we furthermore assume} that the Hilbert spaces are defined in such a way that $\lambda_{\cdot,d}$ is non-decreasing for every $d$, as spelled out in \eqref{eq:eigenvalues}, although this assumption is not needed for our results.

@@

\peter{TODO: Maybe add a short paragraph on notation used in the paper and on its structure? In particular, we should introduce 
an ordering of the parameters $\vp$.}

\section{(Strong) Tractability}

@@

\paragraph{Example 1: Strong polynomial tractability}

Let $T: (0,\infty) \times \naturals \times [0,\infty) \times [0,\infty)\Rightarrow [1,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (\max\{\varepsilon^{-1},1\})^p .
\]
Then Theorem \ref{thm_main_strong_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L(p)}^\infty (\min\{\lambda_{i,d},1\})^p < \infty.
\]
This essentially recovers the result on strong polynomial tractability in \cite[Theorem 5.1]{NW08}

\paragraph{Example 2: Polynomial tractability}

Let $T: (0,\infty) \times \naturals \times [0,\infty) \times [0,\infty)\Rightarrow [1,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (\max\{\varepsilon^{-1},1\})^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
This essentially recovers the result on polynomial tractability in \cite[Theorem 5.1]{NW08}

\peter{(PETER: Add further examples (e.g., EXP- polynomial tractability, etc.)}

\fred{We need to check the proof for
\begin{itemize}
\item correctness
\item exponents
\item whether the additional assumptions are too restrictive and can be generalized
\item whether we need $\tau$ in the assumption
\item Do we need non-decreasing or strictly increasing?
\end{itemize}}
\begin{definition}[Weak Tractability]
If $n(\varepsilon,d)$ denotes the minimal number of function values needed to approximate all functions. We say that the problem is \emph{weakly tractable} if
    \[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\ln n(\varepsilon,d)}{\varepsilon^{-1}+d} = 0\]
\end{definition}
\begin{definition}[Quasi-polynomial Tractability]
    The problem is \emph{quasi-polynomially tractable}, if there exist two constants $C, t> 0$ such that
    \[
    n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}
    \] for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.
\end{definition}

@@

\end{proof}
}
\paragraph{Example 3: EXP-strong polynomial tractability}

Let $T: (0,\infty) \times \naturals \times [0,\infty) \times [0,\infty)\Rightarrow [1,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log\max\{\varepsilon^{-1},1\})^p.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p  < \infty.
\]
\paragraph{Example 4: EXP-polynomial tractability}

Let $T: (0,\infty) \times \naturals \times [0,\infty) \times [0,\infty)\Rightarrow [1,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log \max\{\varepsilon^{-1},1\})^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
\paragraph{Example 5: $(t,s)$-Weak tractability}

Let $T: (0,\infty) \times \naturals \times [0,\infty) \times [0,\infty)\Rightarrow [1,\infty)$. Suppose we have $T$-$(s,t)$- weak tractability, such that
\[
 \lim_{d+\varepsilon^{-1}\rightarrow\infty} \frac{\log \max(1,n(\varepsilon,d))}{d^t +(\max(1,\varepsilon^{-1}))^s} =0.
\]
Then Theorem \ref{thm:weak} yields the iff-condition
\[
\sup_{d\in\naturals} \sigma (c,d,s,\vp) \exp\left(-c\, T(0,d^t,\vp)\right)<\infty
\]
where $\sigma (c,d,s,\vp) = $

\section{$T$-Quasi-polynomial tractability} \label{sec:qpt}

\begin{definition}[$T$-Quasi-polynomial tractability]
We say that the problem is \emph{$T$- quasi-polynomial tractable} if there exists positive $C_{\vp}$ such that
   %\[ n(\varepsilon,d)\leq C_{\vp}\exp\{(T(0,d,\vp))(T(\varepsilon^{-1},1,\vp)\}\]
   \[ n(\varepsilon,d)\leq C_{\vp}\exp\{(T(\varepsilon^{-1},d,\vp)\}\]
\end{definition}

\begin{theorem}\label{thm:qpt}
A problem is $T$-quasi-polynomial tractable iff there exists $\vp \in [\vzero, \boldsymbol{\infty})$ such that
\begin{equation}\label{thmqpt}
M_{\vp} := \sup_{d\in\mathbb{N}}\sum_{i=1}^\infty \frac{1}{T(\lambda_{i,d}^{-1},d,\vp)}
<\infty
\end{equation}
%where $\Bar{p} = -p(1+\log(T(0,d,\vp))$.
%If it holds then 
%\[n(\varepsilon,d)\leq 1+T(0,d^\tau,\vp)+ T(0,d^\tau,\vp)T(\varepsilon^{-1},1,\vp)\]
\end{theorem}
\begin{proof}
\textbf{Necessity Condition:}\\
Suppose that we have $T$-quasi-polynomial tractability, i.e., for some $\vp\geq 0$, there exists positive $C_{\vp}$ such that
%\[\comp(\varepsilon,d)\le C_{\vp}\exp\{(T(0,d,\vp)(T(\varepsilon^{-1},1,\vp))\}\quad\quad \forall\varepsilon >0, d\in \mathbb{N}\] 
\[
\comp(\varepsilon,d)\le C_{\vp}\exp\{T(\varepsilon^{-1},d,\vp)\}
\quad\quad \forall\varepsilon >0, d\in \mathbb{N}
\] 
Since $\lambda_{i,d}$ are non-increasing, we have
\[
\lambda_{\left\lfloor C_{\vp}\exp\{T(\varepsilon^{-1},d,\vp)\}\right\rfloor+1,d}\leq \varepsilon^2
\]
Given that the estimate holds for all $\varepsilon> 0$, we assume that $\varepsilon \in (0, 1]$.\\

If we vary $\varepsilon\in (0,1]$, we have that
\[
j(\varepsilon,d,\vp) = \left\lfloor C_{\vp}\exp\{T(\varepsilon^{-1},d,\vp)\}\right\rfloor+1
\] which attains the values $j = j_d,j_{d+1},\dots$, where
\[
j(\infty,d,\vp)=j_d = \lfloor C_{\vp}\exp\{ T(0,d,\vp)\}\rfloor + 1
\]
Furthermore, we have
\[j \leq C_{\vp}\exp\{T(\varepsilon^{-1},d,\vp)\}+1 \leq C_{\vp}\exp\{T(\lambda^{-1}_{j,d},d,\vp)\}+1 \]
This implies that for $\tau>1$, we have
%\[\left(\log\frac{j-1}{C_{\vp}}\right)^\tau \leq \left(T(\varepsilon^{-1},d,\vp)\right)^\tau \leq K_{\vp,\tau}T(\lambda_{j,d},d,\tau\vp) \]
we have 
%\[\frac{1}{T(\lambda_{i,d}^{-1},d,\vp)} \leq \frac{K_{\vp,\tau}}{\left(\log\frac{j-1}{C_{\vp}}\right)^\tau}\]
\end{proof}

\section{Fred's Attempt at sub-$h$ $T$-Tractability}
May we frame weak tractability more generally?

Let $h:[0,\infty) \to [1,\infty)$ satisfy
\begin{equation} \label{eq:h_cond}
	h \text{ is strictly increasing}, \qquad 
	h(0)=1, \qquad  [h(x)]^c \le h(c x) \quad \forall x \ge 0, c > 0.
\end{equation}
\peter{Let us make the slightly stronger assumption on $h$ that
\[
h(x)\cdot h(y) \le h(x+y)\quad \forall x,y\in [0,\infty),
\]
and let us denote the inverse function of $h$ by $h^{-1}$. 
Then, for arbitrary $z,w \in [1,\infty)$ we choose $x=h^{-1}(z)$ and $y=h^{-1}(w)$, to obtain
\[
z w \le h (h^{-1}(z)+h^{-1}(w)),
\]
which yields by the monotonicity of $h^{-1}$ that
\begin{equation}\label{eq:submult_h_inverse}
h^{-1}(z w) \le h^{-1}(z)+h^{-1}(w).
\end{equation}
}

\fred{There is a problem here because I was assuming that 
\[
\lim_{x \to \infty} \frac{f(x)}{g(x)} = 0 \iff 
\lim_{x \to \infty} \frac{\log(f(x))}{\log(g(x))} = 0.
\]
But this is NOT the case, e.g., $f(x) = x^2$ and $g(x) = x^4$.  So we may need 
\begin{definition} \label{def:subhT}
	A problem is sub-$h \circ T$ tractable if
	\begin{equation*}
		\lim_{\varepsilon^{-1} + d \to \infty} \frac{h^{-1}(\comp(\varepsilon,d))}{T(\varepsilon^{-1},d,\vp)} = 0.
	\end{equation*}
\end{definition}
And NOT what I had below.}

\begin{definition} \label{def:subhT}
	A problem is sub-$h \circ T$ tractable if
	\begin{equation*}
		\lim_{\varepsilon^{-1} + d \to \infty} \frac{\comp(\varepsilon,d)}{h(T(\varepsilon^{-1},d,\vp))} = 0.
	\end{equation*}
\end{definition}

@@

\begin{thebibliography}{99}

\bibitem{HKW19} F.J.~Hickernell, P. Kritzer, H.~Wo\'zniakowski.  Exponential tractability of linear tensor product problems. In: 
D.R.~Wood, J.~DeGier, C.~Praeger, T.~Tao (eds.), 2018 MATRIX Annals, 61--78, Springer, Cham, 2020. 
\bibitem{KW19} P. Kritzer, H.~Wo\'zniakowski.
Simple characterizations of exponential tractability for linear
multivariate problems. J.~Complexity 51, 110--128, 2019.
\bibitem{NW08} E.~Novak, H.~Wo\'zniakowski. \textit{Tractability of Multivariate Problems, Volume I: Linear Information}.
European Mathematical Society Publishing House, Zurich, 2008.
 \bibitem{WW17} A.G.~Werschulz, H.~Wo\'zniakowski. A new characterization of $(s,t)$-weak tractability. J. Complexity 38, 68--79, 2017. 
\end{thebibliography}

>>>>

\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

The goal is to find an approximate solution, $\APP_d:\cb_d \times (0,\infty) \to \cg_d$ satisfying
\begin{equation}
    \label{eq:errcrit}
    \norm[\cg_d]{\SOL_d(f) - \APP_d(f,\varepsilon)} \le \varepsilon \qquad \forall f \in \cb_d, \ \varepsilon \in (0,\infty)
\end{equation}
where $\cb_d$ is the unit ball in $\cf_d$ and $\APP_d(f,\varepsilon)$ is allowed to depend on arbitrary linear functionals \peter{(note that we also allow adaptive information, i.e., the linear functionals may sequentially depend on each other)}.

The information complexity of the problem $\SOL_d$ is denoted $\comp(\varepsilon,d)$ and defined as the minimum number of linear functionals required to satisfy the error criterion \eqref{eq:errcrit}, 
\begin{equation*}
\comp:(0,\infty) \times \naturals \to \natzero.
\end{equation*}
By definition, the complexity is non-decreasing  as $\varepsilon$ tends to zero.  We also expect it to be non-decreasing as $d$ tends to infinity.  \peter{This means that we are actually considering 
a whole sequence of problems  $\SOL_d : \cf_d \to \cg_d$, for $d\in\naturals$, and we furthermore assume} that the Hilbert spaces are defined in such a way that $\lambda_{\cdot,d}$ is non-decreasing for every $d$, as spelled out in \eqref{eq:eigenvalues}, although this assumption is not needed for our results.

For this case where arbitrary linear functionals are allowed, the optimal solution is
\[
\APP_d(f,\varepsilon) = \sum_{i=1}^n \SOL(u_i) \ip[\cf_d]{f}{u_i},
\]
and the \peter{information} complexity of our linear problem is
\begin{equation}\label{eq:comp_def}
\comp(\varepsilon, d) = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\}, \qquad \comp:(0,\infty) \times \naturals \to \natzero.
\end{equation}

@@

\peter{TODO: Maybe add a short paragraph on the notation used in the paper and on its structure? In particular, we should introduce 
an ordering of the parameters $\vp$.}\\
\kachi{Throughout the article, we use the following notation.}\\

\kachi{In Section \ref{sec:spt}, we give the definitions of the considered tractability notions, in particular (strong) tractability. We then consider the strong tractability where the complexity is independent of the dimension, $d$. We then turn to tractability. In Section \ref{sec:examples} we provided examples of strong(traceability) functions for both the algebraic and exponential cases. }

\section{(Strong) Tractability}\label{sec:spt}

@@

\begin{example}[Strong polynomial tractability]

Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (\max\{\varepsilon^{-1},1\})^p .
\]
Then Theorem \ref{thm_main_strong_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L(p)}^\infty (\min\{\lambda_{i,d},1\})^p < \infty.
\]
This essentially recovers the result on strong polynomial tractability in \cite[Theorem 5.1]{NW08}
\end{example}

\begin{example}[Polynomial traceability]
Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (\max\{\varepsilon^{-1},1\})^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
This essentially recovers the result on polynomial tractability in \cite[Theorem 5.1]{NW08}
\end{example}

\peter{(PETER: Add further examples (e.g., EXP- polynomial tractability, etc.)}
\kachi{\begin{example}[EXP-strong polynomial tractability]
Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log\max\{\varepsilon^{-1},1\})^p.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p  < \infty.
\]
\end{example}
\begin{example}[EXP-polynomial traceability]

Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= (1+\log \max\{\varepsilon^{-1},1\})^p\, d^q.
\]
Then Theorem \ref{thm_main_tract2} yields the iff-condition
\[
 \sup_{d\in\naturals} \sum_{i=L_{\vp} d^{q_1}}^\infty (\min\{\lambda_{i,d},1\})^p d^{-q} < \infty.
\]
\end{example}}
\bigskip

\fred{We need to check the proof for
\begin{itemize}
\item correctness
\item exponents
\item whether the additional assumptions are too restrictive and can be generalized
\item whether we need $\tau$ in the assumption
\item Do we need non-decreasing or strictly increasing?
\end{itemize}}
\begin{definition}[Weak Tractability]
If $n(\varepsilon,d)$ denotes the minimal number of function values needed to approximate all functions. We say that the problem is \emph{weakly tractable} if
    \[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\ln n(\varepsilon,d)}{\varepsilon^{-1}+d} = 0\]
\end{definition}
%\begin{definition}[Quasi-polynomial Tractability]
 %   The problem is \emph{quasi-polynomially tractable}, if there exist two constants $C, t> 0$ such that
  %  \[
   % n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}
    %\] for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.
%\end{definition}

@@

\end{proof}
}

%\paragraph{Example 5: $(t,s)$-Weak tractability}
%Let $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$. Suppose we have $T$-$(s,t)$- weak tractability, such that
%\[ \lim_{d+\varepsilon^{-1}\rightarrow\infty} \frac{\log \max(1,n(\varepsilon,d))}{d^t +(\max(1,\varepsilon^{-1}))^s} =0.\]
%Then Theorem \ref{thm:weak} yields the iff-condition
%\[
%\sup_{d\in\naturals} \sigma (c,d,s,\vp) \exp\left(-c\, T(0,d^t,\vp)\right)<\infty
%\]
%where $\displaystyle\sigma (c,d,s,\vp) =\sum_{i=1}^\infty \exp\left(-c\left(\lambda^{-1}_{i,d}\right)^{s/2}\right) $
\kachi{
\begin{example}[Quasi-$T$-polynomial tractability] 
\begin{definition}
    As defined in \cite{GW11}, we say that a problem is \emph{quasi-polynomially tractable} if there exist two constants $C$, $t > 0$, such that
\[
n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}\] 
for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.\\
\end{definition}
Suppose $ T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty)$
be defined by
\[
 T(\varepsilon^{-1},d,\vp)= \exp\{\vp(1+\ln d)(1+\ln \max(\varepsilon^{-1},1))\}
\]
\end{example}

Then Theorem \ref{sec:tractability} yields the iff-condition as follows

\[
\sup_{d\in\mathbb{N}}\sum_{i=Lpd^{q_1}} \left(\min\{\lambda_{i,d},1\} \right)^{p(1+\ln d)}d^{-q} < \infty
\]
}

\section{Sub-$h$ Tractability}
In this section we generalize the concept of weak tractability \cite{}.  Let  $h:[0,\infty) \to [1,\infty)$ satisfy
\begin{equation} \label{eq:h_cond}
	h \text{ is strictly increasing}, \qquad 
	h(0)=1, \qquad  
 h(x+y) \ge  h(x)\cdot h(y)  \quad \forall x,y \ge 0.
\end{equation}
This means that $\log(h(\cdot))$ is a superadditive function.  The function $h$ here can be thought of as the exponential function for weak tractability.

Furthermore, let us denote the inverse function of $h$ by $h^{-1}$. 
Then, for arbitrary $z,w \in [1,\infty)$ we choose $x=h^{-1}(z)$ and $y=h^{-1}(w)$, to obtain
\[
z w \le h (h^{-1}(z)+h^{-1}(w)),
\]
which yields by the monotonicity of $h^{-1}$ that
\begin{equation}\label{eq:submult_h_inverse}
h^{-1}(z w) \le h^{-1}(z)+h^{-1}(w).
\end{equation}
Now we define a generalized notion of weak tractability.
\begin{definition} \label{def:subhT}
	A problem is sub-$h \circ T$ tractable if
	\begin{equation*}
		\lim_{\varepsilon^{-1} + d \to \infty} \frac{h^{-1}(\max(1,\comp(\varepsilon,d)))}{T(\varepsilon^{-1},d,\vp)} = 0.
	\end{equation*}
\end{definition}

\fred{Can we say that this notion of tractabilty is stronger or weaker than $\comp(\varepsilon,d)/h(T(\varepsilon^{-1},d,\vp)) \to 0$ as $\varepsilon^{-1} + d \to \infty$?}

@@

\begin{thebibliography}{99}
\bibitem{GW11} M. Gnewuch, H.~~Wo\'zniakowski. Quasi-polynomial tractability, J.~Complexity, 27, 312--330 2011.
\bibitem{HKW19} F.J.~Hickernell, P. Kritzer, H.~Wo\'zniakowski.  Exponential tractability of linear tensor product problems. In: 
D.R.~Wood, J.~DeGier, C.~Praeger, T.~Tao (eds.), 2018 MATRIX Annals, 61--78, Springer, Cham, 2020. 
\bibitem{KW19} P. Kritzer, H.~Wo\'zniakowski.
Simple characterizations of exponential tractability for linear
multivariate problems. J.~Complexity 51, 110--128, 2019.
\bibitem{NW08} E.~Novak, H.~Wo\'zniakowski. \textit{Tractability of Multivariate Problems, Volume I: Linear Information}.
European Mathematical Society Publishing House, Zurich, 2008.
 \bibitem{WW17} A.G.~Werschulz, H.~Wo\'zniakowski. A new characterization of $(s,t)$-weak tractability. J. Complexity 38, 68--79, 2017.

\end{thebibliography}
